
This material was prepared as an account of work sponsored by an agency of the
United States Government.  Neither the United States Government nor the United
States Department of Energy, nor Battelle, nor any of their employees, MAKES
ANY WARRANTY, EXPRESS OR IMPLIED, OR ASSUMES ANY LEGAL LIABILITY OR
RESPONSIBILITY FOR THE ACCURACY, COMPLETENESS, OR USEFULNESS OF ANY
INFORMATION, APPARATUS, PRODUCT, SOFTWARE, OR PROCESS DISCLOSED, OR REPRESENTS
THAT ITS USE WOULD NOT INFRINGE PRIVATELY OWNED RIGHTS.


\begin{center}
{\bf LIMITED USE}
\end{center}

This software (including any documentation) is being made available to
you for your internal use only, solely for use in performance of work
directly for the U.S. Federal Government or work under contracts with
the U.S. Department of Energy or other U.S. Federal Government
agencies.  This software is a version which has not yet been evaluated
and cleared for commercialization.  Adherence to this notice may be
necessary for the author, Battelle Memorial Institute, to successfully
assert copyright in and commercialize this software.  This software is
not intended for duplication or distribution to third parties without
the permission of the Manager of Software Products at Pacific
Northwest National Laboratory, Richland, Washington, 99352.

\begin{center}
{\bf ACKNOWLEDGMENT}
\end{center}

This software and its documentation were produced with Government support under
Contract Number DE-AC06-76RLO-1830 awarded by the United States Department of
Energy.  The Government retains a paid-up non-exclusive, irrevocable worldwide
license to reproduce, prepare derivative works, perform publicly and display
publicly by or for the Government, including the right to distribute to other
Government contractors.


\clearpage

\begin{center}
{\bf AUTHOR DISCLAIMER}
\end{center}

This software contains proprietary information of the authors, Pacific
Northwest National Laboratory (PNNL), and the US Department of Energy (USDOE).
The information herein shall not be disclosed to others, and shall not
be reproduced whole or in part, without written permission from PNNL or
USDOE.  The information contained in this document is provided ``AS
IS'' without guarantee of accuracy.  Use of this software is
prohibited without {\bf written} permission from PNNL or USDOE.  The
authors, PNNL, and USDOE make no representations or warranties
whatsoever with respect to this software, including the implied
warranty of merchant-ability or fitness for a particular purpose.  The
user assumes all risks, including consequential loss or damage, in
respect to the use of the software.  In addition, PNNL and the authors
shall not be obligated to correct or maintain the program, or notify
the user community of modifications or updates that will be made over
the course of time.
