% $Id$
\label{sec:interface}

NWChem has interfaces to several different packages which are listed below.
In general, the NWChem authors work with the authors of the other packages
to make sure that the interface works.  However, any problems with the 
interface should be reported to the
{\tt nwchem-users@emsl.pnl.gov} e-mail list.


\section{{\tt DIRDYVTST} --- DIRect Dynamics for Variational Transition State Theory}
\label{sec:dirdyvtst}

by Bruce C. Garrett,\\
Environmental Molecular Sciences Laboratory,\\
Pacific Northwest Laboratory, Richland, Washington\\
\\
Yao-Yuan Chuang and Donald G. Truhlar,\\
Department of Chemistry and Super Computer Institute,\\
University of Minnesota, MN 55455-0431\\
\\
and interfaced to NWChem by\\
\\
Ricky A. Kendall,\\
Scalable Computing Laboratory,\\
Ames Laboratory and Iowa State University, Ames, IA 50011\\
\\
Theresa L. Windus,\\
Environmental Molecular Sciences Laboratory,\\
Pacific Northwest Laboratory, Richland, Washington

If you use the DIRDYVTST portion of NWChem, please use following citation
in addition to the usual NWChem citation from Section \ref{sec:intro}:
\begin{quote}
  DIRDYVTST, Yao-Yuan Chuang and Donald G. Truhlar,
  Department of Chemistry and Super Computer Institute,
  University of Minnesota; Ricky A. Kendall,Scalable Computing Laboratory,
  Ames Laboratory and Iowa State University; Bruce C. Garrett and Theresa L.
  Windus, Environmental Molecular Sciences Laboratory, Pacific Northwest
  Laboratory.
\end{quote}

\subsection{Introduction}

By using DIRDYVTST, a user can carry out electronic structure calculations
with NWChem and use the resulting energies, gradients, and Hessians for
direct dynamics calculations with \htmladdnormallink{POLYRATE}
{http://comp.chem.umn.edu/polyrate/}.
This program prepares the file30 input for \htmladdnormallink{POLYRATE}
{http://comp.chem.umn.edu/polyrate/} from NWChem
electronic structure calculations of energies, gradients and Hessians at the 
reactant, product, and saddle point geometries and along the minimum
energy path.  Cartesian geometries for the reactants, products, and 
saddle points need to be input to this program; optimization of 
geometries is not performed in this program.  Note that \verb+DIRDYVTST+ is
based on the \htmladdnormallink{DIRDYGAUSS program} 
{http://comp.chem.umn.edu/dirdyg/}
and is similar to two other programs: \htmladdnormallink{DDUTILITIES}
{http://comp.chem.umn.edu/ddutil/} and \htmladdnormallink{GAUSSRATE}
{http://comp.chem.umn.edu/gaussrate/}.  Users of this module are
encouraged to read the 
\htmladdnormallink{POLYRATE}
{http://comp.chem.umn.edu/polyrate/} manual since they will need to
create the file fu5 input to run calculations with \htmladdnormallink{POLYRATE}
{http://comp.chem.umn.edu/polyrate/}.

Notes about the code:

Input.  The code has been written to parallel, as much as possible,
the \verb+POLYRATE+ code.  

Output.  There is one default output file for each \verb+DIRDYVTST+ run - .file30.

Integrators for following the reaction path.  
Currently the Euler and three Page-McIver (PM) methods 
are implemented.  The PM methods are the local quadratic approximation 
(LQA), the corrected LQA (CLQA), and the cubic (CUBE) algorithm.
The PM methods are implemented so that the Hessian can be reused at
intermediate steps at which only the gradient is updated.  


\subsection{Files}

Test runs are located in directories in \verb+$NWCHEM_TOP/QA/tests+.  Test 
runs are available for two systems: $H + H_2$ and $OH + H_2$.

The $H + H_2$ test uses the Euler integration method at the SCF/3-21G level
of theory to calculate points along the reaction path.
This test is located in the \verb+$NWCHEM_TOP/QA/tests/h3tr1+ directory.

The $OH + H_2$ test uses the Page-McIver CUBE algorithm to calculate points
on the SCF/3-21G surface and does additional single point calculations at
the SCF/6-31G* level of theory.  This test is located in the
\verb+$NWCHEM_TOP/QA/tests/oh3tr3+ directory.

Note: These tests are set up with SCF, however, other levels of
theory can be used.  The initial hessian calculations at the reactants,
products and saddle point can cause some problems when numerical hessians
are required (especially when there is symmetry breaking in the wavefunction).

\subsection{Detailed description of the input}

The input consists of keywords for NWChem and keywords related to \verb+POLYRATE+
input.  The first set of inputs are for NWChem with the general input block
of the form:

\begin{verbatim}
  DIRDYVTST [autosym [real tol default 1d-2] | noautosym]
    [THEORY <string theory> [basis <string basis default "ao basis">] \
                         [ecp <string ecp>] [input <string input>]]
    [SPTHEORY <string theory> [basis <string basis default "ao basis">] \
                         [ecp <string ecp>] [input <string input>]]
    ...
  END
\end{verbatim}

\subsubsection{Use of symmetry}
The use of symmetry in the calculation is controlled by the keyword
\verb+autosym | noautosym+ which is used as described in the geometry
directive (see Section \ref{sec:geom}).  
\verb+Autosym+ is on by default.  A couple words of warning here.
The tolerance related to \verb+autosym+ can cause problems when taking the
initial step off of the transition state.  If the tolerance is too large and
the initial step relatively small,
the resulting geometry will be close to a higher symmetry than is really 
wanted and the molecule will be symmetrized into the higher symmetry.
To check this, the code prints out the symmetry at each geometry along the 
path.  It is up to the user to check the symmetry and make sure that
it is the required one.  In preverse cases, the user may need to turn
\verb+autosym+ off (\verb+noautosym+) if changing the tolerance doesn't
produce the desired results.  In the case that autosym is used, the
user does not need to worry about the different alignment of the molecule
between NWChem and POLYRATE, this is taken care of internally in the
DIRDYVTST module.

\subsubsection{Basis specification}
The basis name on the theory or sptheory directive is that
specified on a basis set directive (see Section \ref{sec:basis}) and
{\em not} the name of a standard basis in the library.  If not
specified, the basis set for the sptheory defaults to the
theory basis which defaults to \verb+"ao basis"+.  

\subsubsection{Effective core potentials}
If an effective core potential is specified in the usual fashion (see
Section \ref{sec:ecp}) outside of the \verb+DIRDYVTST+ input then this will be
used in all calculations.  If an alternative ECP name (the name
specified on the ECP directive in the same manner as done for basis
sets) is specified on one of the theory directives, then this ECP will
be used in preference for that level of theory.

\subsubsection{General input strings}

For many purposes, the ability to specify the theory, basis and
effective core potential is adequate.  All of the options for each
theory are determined from their independent input blocks.  However,
if the same theory (e.g., DFT) is to be used with different options
for theory and sptheory, then the general input strings must
be used.  These strings are processed as NWChem input each time the
theoretical calculation is invoked.  The strings may contain any NWChem
input, except for options pertaining to \verb+DIRDYVTST+ and the task directive.
The intent is that the strings be used just to control the options
pertaining to the theory being used.

A word of caution.  Be sure to check that the options are producing
the desired results.  Since the NWChem database is persistent,
the input strings should fully define the calculation you wish to have happen.

For instance, if the theory model is DFT/LDA/3-21g and the
sptheory model is DFT/B3LYP/6-311g**, the \verb+DIRDYVTST+ input might look like this
\begin{verbatim}
    dirdyvtst
      theory  dft basis 3-21g    input "dft\; xc\; end"
      sptheory dft basis 6-311g** input "dft\; xc b3lyp\; end"
      ....
    end
\end{verbatim}
The empty \verb+XC+ directive restores the default LDA
exchange-correlation option (see Section \ref{sec:xc}).  Note that
semi-colons and other quotation marks inside the input string must be
preceded by a backslash to avoid special interpretation.

\subsubsection{POLYRATE related options}

These keyword options are simlar to the \verb+POLYRATE+ input format, except there
are no ENERGETICS, OPTIMIZATION, SECOND, TUNNELING, and RATE sections.

\paragraph{GENERAL section}


The GENERAL section has the following format:

\begin{verbatim}
*GENERAL
  [TITLE <string title>]
  ATOMS
    <integer num> <string tag> [<real mass>]
    ...
  END
  [SINGLEPOINT]
  [SAVEFILE (vecs || hess || spc)
\end{verbatim}

  Descriptions
  
    TITLE is a keyword that allows the user to input a description of
              the calculation.  In this version, the user can only have a 
              single-line description.

              For example: 
                          TITLE Calculation of D + HCl reaction

    ATOMS is a list keyword that is used to input a list of 
              the atoms.  It is similar to \verb+POLYRATE+ in that the order of the 
              atom and the atomic symbol are required in a single line.  If
              isotope of the element is considered then the atomic mass is 
              required in units of amu.

              For example:
\begin{verbatim}
                          ATOMS
                            1     H     2.014
                            2     H 
                            3     Cl
                          END
\end{verbatim}

    SINGLEPOINT is a keyword that specifies that a single
	      point calculation is to be performed at the reactants,
              products and saddle point geometries.  The type of 
              single point calculation is specified in the \verb+sptheory+ 
              line.

    SAVEFILE is a keyword that specifies that NWChem files
	      are to be saved. Allowed values of variable input to
	      SAVEFILE are vecs, hess, and spc for saving the files 
	      base theory movecs, base theory hessian and singlepoint
          calculation movecs.
             
\paragraph{REACT1, REACT2, PROD1, PROD2, and START sections}

These sections have the following format:
\begin{verbatim}
*(REACT1 || REACT2 || PROD1 || PROD2 || START)
  GEOM
    <integer num> <real x y z>
    ...
  END
  SPECIES (ATOMIC || LINRP || NONLINRP || LINTS || NONLINTS default NONLINRP)
\end{verbatim}

  REACT1 and REACT2 are input for each of the reactants and PROD1 and PROD2
  are input for each of the products.  REACT1 and PROD1 are required.  START
  is the input for the transition state if one exists, or starting point to
  follow downhill the MEP.

  Descriptions

    GEOM is a list keyword that indicates the geometry of the molecule
              in Cartesian coordinates with atomic unit.

              For example:
\begin{verbatim}
                          GEOM
                             1      0.0     0.0     0.0
                             2      0.0     0.0     1.5
                          END 
\end{verbatim}

    SPECIES is a variable keyword that indicates the type of the 
              molecule.  Options are: ATOMIC (atomic reactant or product), 
              LINRP (linear reactant or product), NONLINRP
              (nonlinear reactant or product), LINTS (linear transition 
              state), and NONLINTS (nonlinear transition state).

              For example:
                          SPECIES  atomic

\paragraph{PATH section}

The Path section has the format:

\begin{verbatim}
*PATH
  [SCALEMASS <real scalemass default 1.0>]
  [SSTEP <real sstep default 0.01>]
  [SSAVE <real ssave default 0.1>]
  [SHESS <real shess default SSAVE>]
  [SLP <real slp default 1.0>]
  [SLM <real slm default -1.0>]
  [SIGN (REACTANT || PRODUCT default REACTANT)]
  [INTEGRA (EULER || LQA || CLQA || CUBE default EULER)]
  [PRINTFREQ (on || off default off)]
\end{verbatim}

  Descriptions

    SCALEMASS is a variable keyword that indicates the arbitrary
              mass (in amu) used for mass-scaled Cartesian coordinates. 
              This is the variable called mu in published papers.  Normally,
              this is taken as either 1.0 amu or, for bimolecular reactions,
              as the reduced mass of relative translation of the reactants.
      
    SSTEP is a variable keyword that indicates the numerical step
              size (in bohrs) for the gradient grid.  This is the step
              size for following the minimum energy path.

    SSAVE is a variable keyword that indicates the numerical step
              size (in bohrs) for saving the Hessian grid. At each
              save point the potential and its first and second
              derivatives are recalculated and written to the .file30
              file. For example, if SSTEP=0.01 and SSAVE=0.1, then the
              potential information is written to .file30 every 10
              steps along the gradient grid.

    SHESS is a variable keyword that indicates the numerical step
              size (in bohrs) for recomputing the Hessian when using a
              Page-McIver integrator (e.g., LQA, CLQA, or CUBE).  For
              Euler integration SHESS = SSAVE. For intermediate points
              along the gradient grid, the Hessian matrix from the
              last Hessian calculation is reused.  For example, if
              SSTEP=0.01 and SHESS=0.05, then the Hessian matrix is
              recomputed every 5 steps along the gradient grid.

    SLP is a variable keyword that indicates the positive limit of 
              the reaction coordinate (in bohrs).

    SLM is a variable keyword that indicates the negative limit of 
              the reaction coordinate (in bohrs).

    SIGN is a variable keyword used to ensure the conventional definition
              of the sign of s, $s < 0$ for the reactant side and
              $s > 0$ for the product side, is followed.  \verb+PRODUCT+
              should be used if the eigenvector at the saddle point points
              toward the product side and \verb+REACTANT+ if the
              eigenvector points toward the reactant side.

    INTEGRA is a variable keyword that indicates the integration
              method used to follow the reaction path.  Options are: EULER,
              LQA, CLQA, and CUBE.

    PRINTFREQ is a variable keyword that indicates that projected
              frequencies and eigenvectors will be printed along the MEP.

\subsubsection{Restart}

\verb+DIRDYVTST+ calculations should be restarted through the normal NWChem
mechanism (See Section \ref{sec:start}).  The user needs to change the
\verb+start+ directive to a \verb+restart+ directive and get rid of any
information that will overwrite important information in the RTDB.  The
\verb+file.db+ and \verb+file.file30+ need to be available for the
calculation to restart properly.

\subsubsection{Example}

This is an example that creates the file30 file for POLYRATE for $H + H_2$.
Note that the multiplicity is that of the entire supermolecule, a doublet.
In this example, the initial energies, gradients, and Hessians are calculated
at the UHF/3-21G level of theory and the singlepoint calculations are 
calculated at the MP2/cc-pVDZ level of theory with a tighter convergence
threshold than the first SCF.

\begin{verbatim}
start h3test

basis
 h library 3-21G
end

basis singlepoint
 h library cc-pVDZ
end

scf
   uhf
   doublet
   thresh 1.0e-6
end

dirdyvtst autosym 0.001
  theory scf input "scf\; uhf\; doublet\; thresh 1.0e-06\; end"
  sptheory mp2 basis singlepoint input \
    "scf\; uhf\; doublet\; thresh 1.0e-07\; end"
*GENERAL
  TITLE
    Test run: H+H2 reaction, Page-McIver CLQA algorithm, no restart

  ATOMS
     1    H
     2    H
     3    H
  END

  SINGLEPOINT

*REACT1
   GEOM
     1  0.0   0.0   0.0
     2  0.0   0.0   1.3886144
   END

   SPECIES LINRP

*REACT2
  GEOM
    3    0.0   0.0 190.3612132
  END

  SPECIES  ATOMIC

*PROD2
  GEOM
    1   0.0   0.0 190.3612132
  END

  SPECIES   ATOMIC

*PROD1

  GEOM
    2  0.0   0.0   1.3886144
    3   0.0   0.0   0.0
  END

  SPECIES  LINRP

*START

  GEOM
    1  0.0   0.0  -1.76531973
    2  0.0   0.0   0.0
    3  0.0   0.0   1.76531973
  END

  SPECIES  LINTS

*PATH
  SSTEP  0.05
  SSAVE  0.05
  SLP    0.50
  SLM   -0.50
  SCALEMASS 0.6718993
  INTEGRA   CLQA
end

task dirdyvtst
\end{verbatim}

