%
% $Id$
%
\label{sec:mepgs}


The MEPGS module performs a search for the two critical points on the
potential energy surface connected to a saddle point of the molecule 
defined by input using the \verb+GEOMETRY+ directive (see Section \ref{sec:geom}). 
The algorithm programmed in MEPGS is a {\bf constrained} trust region quasi-newton optimization and approximate energy Hessian updates.


Optional input for this module is specified within the compound
directive,
\begin{verbatim}
  MEPGS 

    OPTTOL <real opttol default 3e-4>

    EPREC <real eprec default 1e-7>

    STRIDE <real stride default 0.1>

    EVIB <real evib default 1e-4>

    MAXMEP <integer maxiter default 250>

    MAXITER <integer maxiter default 20>

    INHESS <integer inhess default 2>

    (BACKWARD || FORWARD) <string default both>

    (MSWG || NOMSWG) <string default NOMSWG>

    (XYZ || NOXYZ) 

  END
\end{verbatim}

\sloppy

\section{Convergence criteria}

The user may request a specific value for the tolerance with the keyword \verb+OPTTOL+ which will couple all the convergence criteria in the following way:

\begin{verbatim}
         GRMS   1.0*OPTTOL
         GMAX   1.5*OPTTOL
         XRMS   4.0*OPTTOL
         XMAX   6.0*OPTTOL
\end{verbatim}

\section{Available precision}

\begin{verbatim}
    EPREC <real eprec default 1e-7>
\end{verbatim}

In performing a constrained trust region optimization the precision of the energy is coupled to the convergence criteria (see Section \ref{sec:tropt:eprec}).
Note that the default EPREC for DFT calculations is 5e-6 instead of 1e-7.

\section{Controlling the step length}

\begin{verbatim}
    STRIDE <real stride default 0.1>
\end{verbatim}

A dynamic stride (\verb+stride+) is used to control the step length during 
the minimum energy path walking when taking the Euler step as starting point.  

\section{Moving away from the saddle point}

\begin{verbatim}
    EVIB <real evib default 1e-4>
\end{verbatim}

The expected decrease in energy (\verb+evib+) assuming a quadratic approximation around the saddle structure to be obtained. 

\section{Maximum number of MEPGS steps}

\begin{verbatim}
    MAXMEP <integer maxmep default 250>
\end{verbatim}

By default at most 250 minimum energy path steps will be taken,
but this may be modified with this directive.

\section{Maximum number of steps}

\begin{verbatim}
    MAXITER <integer maxiter default 20>
\end{verbatim}

By default at most 20 {\bf constrained} geometry optimization steps will be taken,
but this may be modified with this directive. 

\section{Initial Hessian}
\begin{verbatim}
    INHESS <integer inhess default 2>
\end{verbatim}

With this option the MEPGS module will be able to transform
Cartesian Hessian from previous frequency calculation.

\section{Selecting the side to traverse}
\begin{verbatim}
    (BACKWARD || FORWARD) <string default both>
\end{verbatim}

With this option the MEPGS module will select which side of the 
minimum energy path to explore. By default both sides are explored
for a MEPGS run.

\section{Using mass}
\begin{verbatim}
    (MSWG || NOMSWG) <string default NOMSWG>
\end{verbatim}

With this option the MEPGS will trigger the use of mass when following the 
minimum energy path. Mass is not used as default, if mass is used then 
this formally becomes an intrinsic reaction coordinate.  

\section{Minimum energy path saved XYZ file}

\begin{verbatim}
    XYZ [<string xyz default $fileprefix>]
    NOXYZ
\end{verbatim}

The \verb+XYZ+ directive causes the geometry at each calculated 
structure on the minimum energy path to be output into file in the 
permanent directory in XYZ format. 
The optional string will
prefix the filename.  The \verb+NOXYZ+ directive turns this off.

For example, the input
\begin{verbatim}
    mepgs; xyz ; end
\end{verbatim}
will cause a trajectory file filename.xyz to be created
in the permanent directory.  

\section{MEPGS usage}

\begin{verbatim}
    start somename
    geometry; <saddle point body > ; end   
    task theory freq
    freq; reuse somename.hess ; end
    mepgs; <mepgs options> ; end
    task theory mepgs
\end{verbatim}

In the above example, after performing a frequency analysis for the saddle 
point, the information of the force constant matrix is reused (freq directive) 
in order to be able to follow the transition state mode.

\fussy


